---
title: "covid-19"
output:
  pdf_document: default
  html_document: default
author: "Shing"  
date: "2024-04-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Week 3 Lecture - John Hopkin's Covid-19 data set and Week 5 - final assignment

## PART A - Repeat Instructor's Code

Part A is repeat Dr Wall's code that she presented in the video lectures.

Part B is my further wrok to look into covid cases in Hong Kong and Global model of deaths per thousand vs cases per thousand

(you can go to Part B directly for grading of the Week 5 final assignment).

## Load libraries

Load libraries that will be use in this analysis

```{r load library}
library(tidyverse)
library(lubridate)
```

## Load data

Load csv data from John Hopkins github site

```{r load-covid-19}
url_in<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names<-c("time_series_covid19_confirmed_global.csv",
              "time_series_covid19_deaths_global.csv",
              "time_series_covid19_confirmed_US.csv",
              "time_series_covid19_deaths_US.csv" )

urls<-str_c(url_in,file_names)

global_cases<-read_csv(urls[1])
global_deaths<-read_csv(urls[2])
US_cases<-read_csv(urls[3])
US_deaths<-read_csv(urls[4])

```

## Pivot global table

Pivot table by date on no. of cases and no. of deaths for global tables

```{r pivot global}
global_cases <- global_cases %>%
	pivot_longer(col = -c('Province/State',
                              'Country/Region',Lat,Long),
                     names_to ="date",
                     values_to="cases") %>%
        select(-c(Lat,Long))	




global_deaths <- global_deaths %>%
	pivot_longer(col = -c('Province/State',
                              'Country/Region',Lat,Long),
                     names_to ="date",
                     values_to="deaths") %>%
        select(-c(Lat,Long))	



```

## Joint no. of cases and no. of deaths for global

Filter out zero cases after table join

```{r join global cases and deaths}

global <- global_cases %>%
     full_join(global_deaths) %>%
     rename(Country_Region='Country/Region',
            Province_State='Province/State') %>%
     mutate(date=mdy(date))

global <- global %>% filter(cases>0)

```

## Pivot US table

Pivot table by date on no. of cases and no. of deaths for US tables

```{r pivot US}
US_cases <-	US_cases %>%
		pivot_longer(cols = -(UID:Combined_Key),
			names_to="date",
			values_to="cases") %>%
		select(Admin2:cases) %>%
		mutate(date=mdy(date)) %>%
		select (-c(Lat,Long_))




US_deaths <-	US_deaths %>%
		pivot_longer(cols = -(UID:Population),
			names_to="date",
			values_to="deaths") %>%
		select(Admin2:deaths) %>%
		mutate(date=mdy(date)) %>%
		select (-c(Lat,Long_))


```

## Joint no. of cases and no. of deaths for US

```{r join US cases and deaths}
US <- US_cases %>%
  full_join(US_deaths)


```

## Create column Combined_Key in global table

Create column Combined_Key based on Province_State and Country_Region in global table to mimic the table structure of US table

```{r add Combined_Key column to global}

global <- global %>%
	unite("Combined_Key",
		c(Province_State,Country_Region),
		sep =" ,",
		na.rm =TRUE,
		remove=FALSE)

```

## Load Population data

There is no pupulation data in global table, load population data from John Hopkin's git hub site for further analysis

```{r read in population from uid lookup}

uid_lookup_url <-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid<-read_csv(uid_lookup_url) %>%
	select(-c(Lat, Long_,Combined_Key,code3,iso2,iso3,Admin2))
```

## Joint Population data to global table

Join the population data by uid to the global table, remove UID, FIPS column from global table after join operation

```{r join population from uid lookup to global}

global <-global %>%
	left_join(uid,by =c("Province_State","Country_Region")) %>%
	select(-c(UID,FIPS)) %>%
	select(Province_State,Country_Region,date,cases,deaths,Population,Combined_Key)

```

## Calculate death per millon by state in US table

Calculate death per millon by date by state save result to US_by_state

```{r calc US deaths per mill by state per date}
US_by_state <- US %>%
	group_by(Province_State,Country_Region,date) %>%
	summarize(cases =sum(cases),deaths=sum(deaths),Population=sum(Population)) %>%
	mutate(deaths_per_mill = deaths* 1000000/Population) %>%
	select(Province_State, Country_Region,date,cases,deaths,deaths_per_mill,Population) %>%
	ungroup()
```

## Calculate death per millon of US (whole country)

Calculate death per millon by date of US (as a whole country) by summing up total for all US States save result to US_total

```{r calc summary of US_total by summing up US_by_state all by Country+_region}

US_totals <- US_by_state %>%
	group_by(Country_Region,date) %>%
	summarize(cases =sum(cases),deaths=sum(deaths),Population=sum(Population)) %>%
	mutate(deaths_per_mill = deaths* 1000000/Population) %>%
	select(Country_Region,date,cases,deaths,deaths_per_mill,Population) %>%
	ungroup()

```

## Plot cases and deaths of US

Plot cases and deaths from US_totals

```{r plot cases and deaths of US_totals}
US_totals  %>%
	filter(cases >0) %>%
	ggplot(aes(x=date,y=cases)) +
	geom_line(aes(color="cases")) +
	geom_point(aes(color="cases")) +
	geom_line(aes(y=deaths, color="deaths")) +
	geom_point(aes(y=deaths, color="deaths")) +
	scale_y_log10()+
	theme(legend.position="bottom",
	      axis.text.x=element_text(angle=90))+
	labs(title="COVID19 in US",y=NULL)
```

## Plot cases and deaths of New York State

Plot cases and deaths of New York State from US_by_state

```{r plot cases and deaths of New York using US_by_state}
state <- "New York"


#Save the plot for Part B
NY_case_plot<-
US_by_state  %>%
	filter(Province_State==state) %>%
	filter(cases >0) %>%
	ggplot(aes(x=date,y=cases)) +
	geom_line(aes(color="cases")) +
	geom_point(aes(color="cases")) +
	geom_line(aes(y=deaths, color="deaths")) +
	geom_point(aes(y=deaths, color="deaths")) +
	scale_y_log10()+
	theme(legend.position="bottom",
	      axis.text.x=element_text(angle=90))+
	labs(title=str_c("COVID19 in ",state),y=NULL)


NY_case_plot

```

## Calculate new case and new deaths per day for States of US and US as a whole

Calculate new case and new deaths per day for States of US and US as a whole using the lag() function, save result as new columns in US_by_state and US_totals

```{r calculate new cases/deaths of each date and append the columns to US_by_state and US_totals }

US_by_state <- US_by_state %>%
	mutate(new_cases=cases-lag(cases),
               new_deaths=deaths-lag(deaths))

US_totals <- US_totals %>%
	mutate(new_cases=cases-lag(cases),
               new_deaths=deaths-lag(deaths))

```

## Plot new cases and deaths per day of US (country total)

```{r plot new cases and deaths of US_totals}
US_totals  %>%
	ggplot(aes(x=date,y=new_cases)) +
	geom_line(aes(color="new_cases")) +
	geom_point(aes(color="new_cases")) +
	geom_line(aes(y=new_deaths, color="new_deaths")) +
	geom_point(aes(y=new_deaths, color="new_deaths")) +
	scale_y_log10()+
	theme(legend.position="bottom",
	      axis.text.x=element_text(angle=90))+
	labs(title="New COVID19 cases per day in US",y=NULL)
```

## Plot new cases and deaths per day of New York

```{r plot new cases and deaths select US state}

state <- "New York"

#Save the plot for Part B
NY_new_case_per_date_plot<-
US_by_state  %>%
	filter(Province_State==state) %>%
	ggplot(aes(x=date,y=new_cases)) +
	geom_line(aes(color="new_cases")) +
	geom_point(aes(color="new_cases")) +
	geom_line(aes(y=new_deaths, color="new_deaths")) +
	geom_point(aes(y=new_deaths, color="new_deaths")) +
	scale_y_log10()+
	theme(legend.position="bottom",
	      axis.text.x=element_text(angle=90))+
	labs(title=str_c("New COVID19 cases per day in ",state),y=NULL)

NY_new_case_per_date_plot
```

## Find the 10 States of min and max death per thousand

Find the 10 States of min and max death per thousand (two steps) 1)Calculate min, max, case per thousand, deaths per thousand group by state 2)Use the slice_min(),slice_max() to find the top 10 max and min deaths per thousands

```{r find the 10 States of min and max death per thousand}
US_state_totals <- US_by_state %>%
	group_by(Province_State) %>%
	summarize(deaths =max(deaths),cases =max(cases),population =max(Population),
		  cases_per_thou = 1000*cases/population,
		  deaths_per_thou = 1000*deaths/population) %>%
	filter(cases >0, population>0)



US_state_totals %>% 
slice_min(deaths_per_thou,n=10) %>%
select(deaths_per_thou,cases_per_thou,everything())



US_state_totals %>% 
slice_max(deaths_per_thou,n=10) %>%
select(deaths_per_thou,cases_per_thou,everything())

```

## Create a model

Create a linear model of deaths_per_thou based on cases_per_thou of US State and plot result prediction against actual

```{r create a linear model of deaths_per_thou based on cases_per_thou of US State and plot result prediction against actual}

mod <- lm(deaths_per_thou ~ cases_per_thou, data=US_state_totals)

US_tot_w_pred <- US_state_totals %>% mutate(pred=predict(mod))

US_tot_w_pred %>% ggplot() +
geom_point(aes(x=cases_per_thou,y=deaths_per_thou),color="blue")+
geom_point(aes(x=cases_per_thou,y=pred),color="red")
```

## PART B - My exploration of Hong Kong Data for the final assignment (final assignment)

I would like to explore the data of Hong Kong for the final assignment. Data has already been read and wrangled into global table in Part A, filter out data of Hong Kong and save in Hong_Kong_cases

```{R filter out data of Hong Kong from global and save in Hong_Kong_cases}
Hong_Kong_cases <-global %>%
     filter(Combined_Key=='Hong Kong ,China') %>%
     select(c(Combined_Key, date, cases,deaths,Population))
```

## Plot no. of cases and deaths in Hong Kong

When compare Hong Kong with NY (in Part A), It is observed there is only a single surge in no. of cases in 2020 Spring and the increase in 2021 and 2022 Spring is more gentle. For Hong Kong, the 1st surge in 2020 Spring is not as sharp as NY, however, there is a sharp increase in 2022 Spring, which is not the case in NY. It is worth to look into it why there is the difference, is it due to difference in social distancing order or how the covid vaccine is distributed.

Looking back in time, it seems NY's has a better strategy to cope with the virus, as there is no sharp rise in case and depth in 2022, "normal living" can be resumed earlier in NY.

\*\*\* BIAS WARNING \*\*\*

The above conclusion that NY's has a better strategy may be a ***biased*** view as NY data and Hong Kong data is collected by their respective health agency, which they may have different classification standard/mechanism on whether a case and cause of death is Covid case or not. Therefore, concluding a place's strategy is better than another is ***biased*** by just based on how no. of case is increased in these two places.

```{R plot no. of cases and deaths in Hong Kong, warning=FALSE}
Hong_Kong_cases  %>%
	filter(cases >0) %>%
	ggplot(aes(x=date,y=cases)) +
	geom_line(aes(color="cases")) +
	geom_point(aes(color="cases")) +
	geom_line(aes(y=deaths, color="deaths")) +
	geom_point(aes(y=deaths, color="deaths")) +
	scale_y_log10()+
	theme(legend.position="bottom",
	      axis.text.x=element_text(angle=90))+
	labs(title="COVID19 in Hong Kong",y=NULL)


#Plot the NY cases plot again
NY_case_plot

```

## Calculate new case/death per day and plot graph of Hong Kong

Calculate new case/death pert day and plot new case/death per day of Hong Kong

Comparing to the same graph of NY in Part A, there is more variation in no. of deaths in NY after mid-2022, there is more days with zero death in NY after mid-2022. Hong Kong can only have zero death day in late 2022 but the drop in no. of death is more consistent (less variation) after 2023, while the death per day in NY remain vary a lot after 2023.

```{R calc new cases and death per day, warning=FALSE}
Hong_Kong_cases <- Hong_Kong_cases %>%
	mutate(new_cases=cases-lag(cases),
        new_deaths=deaths-lag(deaths))

Hong_Kong_cases  %>%
	ggplot(aes(x=date,y=new_cases)) +
	geom_line(aes(color="new_cases")) +
	geom_point(aes(color="new_cases")) +
	geom_line(aes(y=new_deaths, color="new_deaths")) +
	geom_point(aes(y=new_deaths, color="new_deaths")) +
	scale_y_log10()+
	theme(legend.position="bottom",
	      axis.text.x=element_text(angle=90))+
	labs(title="New COVID19 cases per day in Hong Kong",y=NULL)

#Plot the NY NY nase case per date plot again
NY_new_case_per_date_plot

```

## Create a simple model for global deaths per thousand based on case per thousand

Steps

1)  Get total no of case and deaths, Population from global group by Country Region

-   It is found some Country Region's Population is null, these Country Region needed is filtered out
-   Calc cases per thousand and deaths per thousand group by Country Region

2)  Create linear model of deaths per thoursand based on case per thousand using lm() of R

3)  Predict deaths per thousand using the linear model, save perdition to global_cases_w_pred

4)  Plot predict value overlay with actual data

```{R}
# Get total no of case and deaths, Population from global group by Country Region
# * It is found some Country Region's Population is null, these Country Region are filtered out before doing the  summation
# * Calc cases per thousand and deaths per thousand group by Country Region

global_totals <- global %>%
  filter(Population>=0) %>%
	group_by(Country_Region) %>%
  summarize(deaths =max(deaths),cases =max(cases),population =max(Population)) %>%
	group_by(Country_Region) %>%
  summarize(cases_per_thou = 1000*cases/population,
		        deaths_per_thou = 1000*deaths/population)


# Create linear model of deaths per thoursand based on case per thousand using lm() of R
global_mod <- lm(deaths_per_thou ~ cases_per_thou, data=global_totals)

# Predict deaths per thousand using the linear model, save perdition to global_cases_w_pred
global_cases_w_pred <- global_totals %>% mutate(pred=predict(global_mod))

# Plot predict value overlay with actual data
global_cases_w_pred %>% ggplot() +
geom_point(aes(x=cases_per_thou,y=deaths_per_thou),color="blue")+
geom_point(aes(x=cases_per_thou,y=pred),color="red") +
labs(title="Linear model of Global deaths per thousand against cases per thousand",y=NULL)
```
